\Large \bf Cyclone: Warp Speed Replication for Key Value Stores

### 2017-07-23 paper/paper.tex
Updates.

\usepackage[normalem]{ulem}

@@

% Use the following at camera-ready time to suppress page numbers.
% Comment it out when you first submit the paper for review.
%\thispagestyle{empty}

@@

These key value stores however tend to ignore an important component: the write
ahead log. A machine or storage failure leading to total loss or temporary
unavailability of data is unacceptable in many internet scale services where
high availabilty and revenue are often interconnected. Key value stores
therefore usually incorporate support for a write ahead log that if replicated
and kept durable for every appended update provides the necessary high
availability. Unfortunately, the write ahead log is a performance achilles heel
for these systems - eclipsing much of the work on improving the performance of
the LSM component. To illustrate the impact of the write ahead log, consider
Figure~\ref{fig:problem}. The line marked 'Rocksdb' shows the performance of
Rocksdb without the write ahead log. The performance when persisting the write
ahead log without replicating it is shown as the line marked 'rocksdb/WAL'. The
line marked 'rocksdb/3 way rep.' is for simply replicating the log three ways
without persisting it, using RAFT~\cite{raft} running over TCP/IP. Either
persisting every update to the log or replicating it using TCP/IP causes
performance to drop by an order of magnitude (note the log scale on the
x-axis). It is therefore no surprise that deployments of Rocksdb often turn off
the write ahead log~\cite{samza}. On the other side of the spectrum, key value
store research prototypes such as FloDB~\cite{flodb} turns off the write ahead
log to be able to showcase benefits of sophisticated extensions to LSM
datastructures.

@@

Cyclone is a high speed strongly consistent replicated write ahead logging
service specialized for key value stores such as Rocksdb. We describe Cyclone's
system architecture in Section~\ref{sec:sysarch}. Cyclone provides the order of
magnitude better performance required to close the performance gap in
Figure~\ref{fig:problem} both on the storage and network side. Cyclone requires
\emph{a small amount of} directly attached Non Volatile Memory (NVM) on the host
- such as in the form of NVDIMMs~\cite{farm} or directly attached persistent
memory~\cite{pmfs}. Cyclone logs update operations to the key value store and
calls on the key value store to execute the update only when it has been
successfully replicated to majority quorum of replicas. All reads and write to
the key value store are always executed by the leader replica. A client side
stub provided by Cyclone automatically locates and directs read and write
requests to the leader replica, ensuring linearizable semantics from the key
value store. Using a replicated write ahead log to build a linearizable and
highly available key value store is not new [find some examples here]. The
novelty in Cyclone is that it accomplishes this without compromising on the
performance baseline of no logging or replication. Cyclone does this by
systematically addressing the sources of performance degradation in
Figure~\ref{fig:problem}.

@@

A detailed evaluation of Cyclone's replication is provided in the context of
RocksDB - a popular persistent key value store in
Section~\ref{sec:evaluation}. We show that Cyclone achieves performance
transparent replication in that it can provide both a persistent write ahead log
and replication without compromising on RocksDB's performance.

\section{System Architecture}
\label{sec:sysarch}

\section{Storage}
\label{sec:storage}
A fundamental challenge in building an efficient replicated log is dealing with
network overheads. Cyclone's construction is predicated on the observation that
in the common case, protocols such as RAFT simply multicast the same entry
received from a client, from the leader replica to all follower replicas. This
suggests that an efficient way to implement a replicated log is to treat it as a
software packet switching problem. However peristing logs on flash represents an
impediment in that block storage devices are not good candidates to hold packets
being software switched. Our solution is to split the log into two levels. The
upper level is held in directly attached non-volatile memory (NVM)
such as NVDIMMs. The lower level is held on a flash SSD - we term the lower
level log as the flashlog. We periodically drain entries in batches from the NVM
log to the backing flashlog. An important consequence of this two level approach
is that we need only a limited amounts of NVM with the bulk of the log residing
on flash - an arrangement that maps well to the current relative costs of the
two types of memory. We begin by describing the NVM log and how we implement the
RAFT consensus protocol in the packet switching paradigm around it. We then
describe the flashlog and how we drain entries in batches from the NVM level to
the flash level.

\subsection{NVM Log}
The NVM log is maintained as a circular log of fixed sized pointers (a pointer
log) to memory buffers - each of which holds a client request. This arrangement
is shown in Figure~\ref{fig:nvm_log} that also illustrates the advantage of such
an arragement. Adding a level of indirection in the NVM log allows us to
separate the FIFO ordered circular log being manipulated by the RAFT from
variable sized client being managed by the memory allocator of the packet
switching library we use. Another advantage is that it makes recovery from NVM
easy - appends to the circular pointer log are atomic and we can use the pointer
log to recover allocator state i.e. what pieces of NVM are currently in use by
the log.

@@

\subsection{Flash Log}
Cyclone is designed to use only limited amounts of NVM and drains log entries
down into a log placed on a flash SSD - that we term the flashlog. The flashlog
is written out in segments of configurable size (we use 128KB segments). A
segment buffer is prepared in memory (volatile DRAM) using the layout shown in
Figure~\ref{fig:flashlog_page}. We do not allow objects in the flashlog to cross
a 4KB boundary - linking multiple objects together with a special flag encoded
into the size if necessary. We flush log segments out to the log file using
asynchronous direct IO, and therefore we fill one log segment buffer while
keeping IO to another one outstanding to the flash drive. To avoid having to do
a synchronous metadata flush, we preallocate (using the posix\_fallocate call) a
gigabyte worth of zero filled disk pages at the end of the log file whenever we
hit its end.

We move log entries from the head of the NVM log to the flashlog buffers in FIFO
order. The NVM log entry is only actually removed when the IO for the
corresponding flashlog page is complete. This means that during recovery we can
have the same log entry both in the NVM log and the flashlog, a condition that
can be detected by examining the RAFT related information that is embedded as
part of the logged packet.

>>>>

\usepackage[normalem]{ulem}
\usepackage{parskip}

@@

@@

These key value stores however tend to ignore an important component: the write
ahead log. A machine or storage failure leading to total loss or temporary
unavailability of data is unacceptable in services where high availabilty and
revenue are interconnected. Key value stores therefore usually incorporate
support for a write ahead log that if replicated and kept durable for every
appended update provides the necessary high availability. Unfortunately, the
write ahead log is a performance achilles heel for these systems - eclipsing
much of the work on improving the performance of the LSM component. To
illustrate the impact of the write ahead log, consider
Figure~\ref{fig:problem}. The line marked 'Rocksdb' shows the performance of
Rocksdb without the write ahead log. The performance when persisting the write
ahead log without replicating it is shown as the line marked 'rocksdb/WAL'. The
line marked 'rocksdb/3 way rep.' is for simply replicating the log three ways
without persisting it, using RAFT~\cite{raft} running over TCP/IP. Either
persisting every update to the log or replicating it using TCP/IP causes
performance to drop by an order of magnitude (note the log scale on the
x-axis). It is therefore no surprise that deployments of Rocksdb often turn off
the write ahead log~\cite{samza}. On the other side of the spectrum, key value
store research prototypes such as FloDB~\cite{flodb} turns off the write ahead
log to be able to showcase benefits of sophisticated extensions to LSM
datastructures.

@@

Cyclone is a high speed strongly consistent replicated write ahead logging
service specialized for key value stores such as Rocksdb. We describe Cyclone's
system architecture in Section~\ref{sec:sysarch}. Cyclone provides the order of
magnitude better performance required to close the performance gap in
Figure~\ref{fig:problem} both on the storage and network side. Cyclone requires
\emph{a small amount of} directly attached Non Volatile Memory (NVM) on the host
- such as in the form of NVDIMMs~\cite{farm} or directly attached persistent
memory~\cite{pmfs}. Cyclone logs update operations to the key value store and
calls on the key value store to execute the update only when it has been
successfully replicated to majority quorum of replicas. All reads and write to
the key value store are always executed by the leader replica. A client side
stub provided by Cyclone automatically locates and directs read and write
requests to the leader replica, ensuring linearizable semantics from the key
value store. The novelty in Cyclone is that it accomplishes this without
compromising on the performance baseline of no logging or replication. Cyclone
does this by systematically addressing the sources of performance degradation in
Figure~\ref{fig:problem}.

@@

A detailed evaluation of Cyclone's replication is provided in the context of
RocksDB - a popular persistent key value store - in
Section~\ref{sec:evaluation}. We show that Cyclone achieves performance
transparent replication in that it can provide both a persistent write ahead log
and replication without compromising on RocksDB's performance.

\section{System Architecture}
\label{sec:sysarch}
A persistent key value store durably stores key-value mappings accessible through a
simple interface:

\begin{itemize}
\item GET(K): returns value corresponding to key K
\item PUT(K, V): sets the value of key K to value V
\end{itemize}

A key value store typically serves multiple client requests concurrently and is
linearizable~\cite{linearizability}. This means any history of requests and
responses can be rearranged such that responses immediately follow corresponding
requests and if a response precedes a request in the original history it also
does so in the rearranged history. Key value stores such as RocksDB also
support atomic writes to multiple keys and snapshots.

Cyclone integrates with key value stores as both a client and server side
library. Cyclone replicates the key value store across a set of replica
servers. The client side library sends requests to a distinguished leader
replica.  On the server side Cyclone accepts requests from clients and calls
into the key value store. It then returns a response to the client. Before
executing any PUT request, Cyclone appends it to a durable log and replicates
the request to the logs of follower replicas. The request is considered
replicated once acknowledged by a majority of replicas.

Cyclone uses RAFT~\cite{raft} to keep the logs in sync across the replicas. On a
failure of the leader replica a new leader is automatically elected - RAFT
ensures that the new leader has the most up to date logs. The Cyclone client
library automatically locates the new leader on a failover - clients therefore
see only a brief interruption in service. Although read requests (GET
operations) are not logged, they are serviced in FIFO order with the write requests
(PUT operations) currently in Cyclone's log. This ensures linearizability in the
face of failovers.

\section{Storage}
\label{sec:storage}
Storage devices such as flash-based SSDs export a block IO interface. Appending
a small update to the write-ahead log requires a synchronous update of an entire
4KB page. This leads to a high baseline latency at even low load, as shown in
Figure~\ref{fig:problem}.  On the other hand small updates (in the order of 100s
of bytes for both the key and value taken together) are commonplace in many
workloads [find refs]. This problem leads many system designers to
underprovision key-value store shards ensuring that it operates at moderate to
high load, using group-commit to batch additions to the log. They therefore pay
a price in latency to ensure the bandwidth to storage is fully utilized to
prevent it from becoming a bottleneck.

Cyclone adopts a different strategy by making use of novel memory technologies
to obviate the need to make this tradeoff in the first place. We assume a small
amount of non-volatile memory directly addressable by the CPU rather than placed
behind a block oriented IO interface. Log appends are done to the ``NVM log''
placed in this non-volatile memory. While we designed Cyclone to leverage new
non-volatile memory technologies that might be available in the
future~\cite{3dxpoint} current non-volatile memory options such are NVDIMMs
are more expensive that DRAM. Therefore, for Cyclone to be economical
we cannot assume that a sufficient amount of NVM is available to hold the entire
log. The log entries can only be garbage collected when the in-memory component
of the key value store is checkpointed to secondary storage. The growing size of
available DRAM means that also provisioning sufficient NVDIMMs to hold the
corresponding log of updates is not feasible given that NVDIMMs cost at least
twice as much as volatile DRAM. We therefore resort to using a second larger log
level placed on a flash SSD. Entries are drained from the NVM log to this flash
log in conveniently large units that are a multiple of 4KB - the optimum IO unit
for flash SSDs. This ensures we get the best
possible throughput from a flash device without paying the price of synchronous
IO for small key-value pairs. At the same time, the amount of non-volatile
memory required is small enough to not add undue cost of the server.

The flashlog is written out in segments of configurable size (we use 128KB
segments). A segment buffer is prepared in memory (volatile DRAM) using the
layout shown in Figure~\ref{fig:flashlog_page}. We do not allow objects in the
flashlog to cross a 4KB boundary - linking multiple objects together with a
special flag encoded into the size if necessary. We flush log segments out to
the log file using asynchronous direct IO, and therefore we fill one log segment
buffer while keeping IO to another one outstanding to the flash drive. To avoid
having to do a synchronous metadata flush, we preallocate (using the
posix\_fallocate call) a gigabyte worth of zero filled disk pages at the end of
the log file whenever we hit its end.

We move log entries from the head of the NVM log to the flashlog buffers in FIFO
order. The NVM log entry is only actually removed when the IO for the
corresponding flashlog page is complete. This means that during recovery we can
have the same log entry both in the NVM log and the flashlog, a condition that
can be detected by examining the log sequence number that we embed in each log
entry.

The two level log design in Cyclone lays the foundation for improvements to the
network stack that we discuss next - a significant contributor to the overheads
in Figure~\ref{fig:problem}. We use RAFT to replicate the NVM log from the
leader replica to the followers. Each replica \emph{independently} drains
committed entries (in terms of RAFT consensus) from the NVM log to the
flashlog. Block IO therefore is no longer a consideration when optimizing data
movement over the network, unlike systems that need to deal with such
problems~\cite{reflex}. Replication reduces to efficiently sending a block of
data already present in directly attached memory over the network.

The core network operation in replication is to receive a request from the
client at the leader replica and send that exact same request (as a log entry)
to all follower replicas. The leader therefore \emph{multicasts} a received
request packet to follower replicas, a problem well studied in the networking
community when building software packet switches. Software packet switches reach
impressive speeds of millions of packets forwarded per core [find refs], a
number far in excess of the few hundreds of packets we manage in
Figure~\ref{fig:problem}. We therefore proceed by approximating a software
packet switch for the networking component of Cyclone. However software packet
switching does not need to deal with failure - usually requiring only a simple
decision for each packet at the node. RAFT introduces significant
protocol level work for each forwarded packet a challenge that we need to deal
with to approach the desired software packet switching efficiency in the common
case.

\subsection{DPDK}
The Data Plane Development Kit~\cite{dpdk} provides low latency userspace access
to an ethernet NIC, permitting the application to directly send and receive raw
ethernet frames via the transmit and receive queues on the NIC. DPDK is often
used by developers of software packet switches and therefore we also leverage
DPDK as a library for building Cyclone.

DPDK does not by itself provide a TCP stack. This is not a problem. RAFT (and
indeed most consensus protocols) tolerate network losses and reordering by
design due to the need to support asynchronous communication. In addition, most
datacenter networks rarely drop or reorder packets and provide full bisection
bandwith between servers that might serve as Cyclone replicas. We therefore
jettisoned TCP and chose to send raw IP packets encapsulated in ethernet
frames. We currently follow this communication model both for server to server
communication between replicas as well as client server communication for
requests and responses. Although a detailed evaluation is made later, switching
from the kernel TCP/IP stack to DPDK reduces the latency between machines in our
testbed from {\tt 24 us} to as low as {\tt 7 us}, providing a significant boost
to performance.

\subsection{Eliminating Data Movement}

Another key problem we had to solve with DPDK was teaching it about persistent
memory. DPDK allows one to plug in a memory allocator. We required the allocator
to support various features such as allowing multiple entities (the cpu core and
NIC driver) to hold reference to DPDK packets and allocate/free in parallel.

The NVM log is maintained as a circular log of fixed sized pointers (a pointer
log) to memory buffers - each of which holds a client request. This arrangement
is shown in Figure~\ref{fig:nvm_log} that also illustrates the advantage of such
an arragement. Adding a level of indirection in the NVM log allows us to
separate the FIFO ordered circular log being manipulated by the RAFT from
variable sized client being managed by the memory allocator of the packet
switching library we use. Another advantage is that it makes recovery from NVM
easy - appends to the circular pointer log are atomic and we can use the pointer
log to recover allocator state i.e. what pieces of NVM are currently in use by
the log.

@@

\subsection{Flash Log}
